---
title: "Daily Shelter Occupancy 2020 (Toronto Open Data)"
author: "Svitlana Protas"
date: "Compiled on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document: default
  word_document: default
params:
  filename: daily-shelter-occupancy-2020.csv
---

```{r}
source.data <- read.csv(params$filename, stringsAsFactors = FALSE)
head(source.data)
summary(source.data)
```
##Print the structure of your dataset.
```{r}
source.data <- read.csv(params$filename, stringsAsFactors = FALSE)
colnames(source.data)
print(nrow(source.data))
print(ncol(source.data))
```
##List the variables in your dataset
```{r}
source.data <- read.csv(params$filename, stringsAsFactors = FALSE)
str(source.data)
```   
##Print the top 15 rows of your dataset.
```{r}
library(tidyverse)
print(source.data[c(1:15), ])
print(source.data[c(1:15), c("SHELTER_PROVINCE", "SECTOR", "OCCUPANCY", "CAPACITY")])
sample_n(source.data, 15)
```

##Write a user defined function using any of the variables from the data set.
```{r}
occupationalStatus <- function(sctr, rCount=NULL){
  if(!is.null(rCount)){
    # to test on subset
    dt<-source.data[c(1:rCount), c("SHELTER_PROVINCE", "SECTOR", "OCCUPANCY", "CAPACITY")]
  } else {
    dt<-source.data[c("SHELTER_PROVINCE", "SECTOR", "OCCUPANCY", "CAPACITY")]
  } 
  subset(dt, tolower(SECTOR)==tolower(sctr))
}
print(occupationalStatus("women", 50))

#print(occupationalStatus("women"))
#print(occupationalStatus("men"))
#print(occupationalStatus("youth"))
#print(occupationalStatus("co-ed"))
#print(occupationalStatus("families"))

occupationalStatus("women")
occupationalStatus("men")
occupationalStatus("youth")
occupationalStatus("co-ed")
occupationalStatus("families")
```

#Use data manipulation techniques and filter rows based on any logical criteria that exist in your dataset.
```{r}
library(tidyverse)

shelterData<-tibble(source.data)
#View(shelterData)
fData<-filter(shelterData, OCCUPANCY > 500 & tolower(SECTOR)=="families")
select(fData, OCCUPANCY_DATE, SHELTER_NAME:SHELTER_CITY, PROGRAM_NAME:CAPACITY)
```

#Remove missing values in your dataset.
```{r}
print(any(is.na(source.data)))
print(nrow(source.data))

cleannedData <- source.data[complete.cases(source.data),]
nrow(cleannedData)
cleannedData
```

#Identify the dependent & independent variables and use reshaping techniques and create a 
#new data frame by joining those variables from your dataset.
```{r}
united_Data<-source.data %>% unite(SHELTER_FULL_ADDRESS, SHELTER_ADDRESS, SHELTER_CITY, SHELTER_PROVINCE, SHELTER_POSTAL_CODE, sep = ', ')
head(united_Data, 20)
```

#Identify and remove duplicated data in your dataset.
```{r}

#to test
#uniqueData<-source.data$SHELTER_NAME %>% unique()
#uniqueData

#distinctData<-source.data %>% distinct(SHELTER_NAME, .keep_all = FALSE)
#distinctData

tibble(source.data)
unique(source.data)
distinct(source.data, .keep_all = FALSE)

removed<-nrow(source.data) - nrow(unique(source.data))
cat("Rows removed: ", removed)

```

#Reorder multiple rows in descending order
```{r}
arrengedData <-select(source.data, OCCUPANCY_DATE, SHELTER_NAME, OCCUPANCY, CAPACITY) %>% arrange(desc(OCCUPANCY)) %>% filter(OCCUPANCY > CAPACITY)
arrengedData
```

#Rename some of the column names in your dataset.
```{r}
# rename SECTOR column name to SHELTER_SECTOR
result<-source.data %>% filter(OCCUPANCY > CAPACITY) %>% rename(SHELTER_SECTOR = SECTOR) %>% 
        select(OCCUPANCY_DATE, SHELTER_NAME, SHELTER_SECTOR, OCCUPANCY, CAPACITY) 
result

result<-source.data %>% subset(OCCUPANCY < 100) %>% select(OCCUPANCY_DATE, SHELTER_NAME, PROGRAM_NAME:CAPACITY)
result
```

#Add new variables in your data frame by using a mathematical function (for e.g. – multiply 
#an existing column by 2 and add it as a new variable to your data frame)
```{r}
source.data[["SpotsAvailabylity"]]<-(source.data$CAPACITY - source.data$OCCUPANCY)

subset(source.data, source.data$SpotsAvailabylity < 0) #Go over Capacity

#show occupacy percentage
source.data<-mutate(source.data, occupayPercentage = source.data$OCCUPANCY*100/source.data$CAPACITY)
sample_n(source.data, 50)
```

#Print the summary statistics of your dataset.
```{r}
tibble(source.data)
summary(source.data)
```

##Create a training set using random number generator engine.
```{r}

#dataset is large, create sample 10% to test
smp<-sample_frac(source.data, 0.01)
smp

#or using number generator engine
dataSample<-bind_rows(
  col1=runif(10, min=0, max=100), 
  col2=runif(10, min=0, max=100), 
  col3 = runif(10, min=0, max=100),
  col4 = runif(10, min=0, max=100),
  col5 = runif(10, min=0, max=100),
  col6 = runif(10, min=0, max=100),
  col7 = runif(10, min=0, max=100),
  col8 = runif(10, min=0, max=100),
  col9 = runif(10, min=0, max=100),
  col10 = runif(10, min=0, max=100)
  )

df<-data.frame(dataSample)
df
```

##Use any of the numerical variables from the dataset and perform the following statistical functions.
```{r}

tibble(source.data)
summary(source.data)

#Mean
avgData<- source.data %>% select(OCCUPANCY_DATE, SECTOR, OCCUPANCY) %>% group_by(OCCUPANCY_DATE, SECTOR) %>% summarize(mean(OCCUPANCY))
avgData

#Median
medianData<- source.data %>% select(OCCUPANCY_DATE, SECTOR, OCCUPANCY) %>% group_by(OCCUPANCY_DATE, SECTOR) %>% summarize(median(OCCUPANCY)) 
medianData

#Mode
mode <- function(numbers){
  which.max(tabulate(numbers))
}

modeData<- source.data %>% select(OCCUPANCY_DATE, SECTOR, OCCUPANCY) %>% group_by(OCCUPANCY_DATE, SECTOR) %>% summarize(mode(OCCUPANCY))
modeData

#Range
rangeData<-source.data %>% select(OCCUPANCY_DATE, SECTOR, OCCUPANCY) %>% group_by(OCCUPANCY_DATE, SECTOR) %>% summarize(range(OCCUPANCY))
rangeData

```

##Plot a scatter plot for any 2 variables in your dataset
```{r}
library(tidyverse)
library(lubridate)
source.data <- read.csv(params$filename, stringsAsFactors = FALSE)

sdT<-as_tibble(source.data)

grpByMonth <- sdT %>% select(OCCUPANCY_DATE, SECTOR, OCCUPANCY) %>% mutate(OCCUPANCY_DATE = months.Date(as.Date(OCCUPANCY_DATE))) %>%
      group_by(OCCUPANCY_DATE, SECTOR) %>% summarise(OCCUPANCY, list(OCCUPANCY=sum))
grpByMonth

sc<-ggplot(grpByMonth, aes(OCCUPANCY_DATE, OCCUPANCY)) + geom_point(aes(color=SECTOR), sixe=3, na.rm=TRUE) + facet_grid(SECTOR ~.) + stat_smooth()+theme_minimal() +
  labs(x="Month", y="Occupancy", title="2020 Shelter Occupancy by Month and Sector") +
  theme(axis.text.x = element_text(angle=45)) +
  scale_y_continuous(labels = scales::comma)
sc

```

##Plot a bar plot for any 2 variables in your dataset 
```{r}
library(tidyverse)
library(lubridate)
source.data <- read.csv(params$filename, stringsAsFactors = FALSE)

sdT<-tibble(source.data)
sdT

bar<-ggplot(sdT)
bar<-ggplot(sdT, aes(x=SHELTER_CITY, y=OCCUPANCY), na.rm=TRUE)

bar + geom_bar(stat='identity', aes(fill=SECTOR), size=1) + theme_light() + 
  labs(y="Shelter Occupancy", x="City Area", title="2020 Shelter Occupancy by Sector") +
  theme(axis.text.x = element_text(angle=45)) + 
  scale_y_continuous(labels = scales::comma)


```

##Find the correlation between any 2 variables by applying least square linear regression model.
```{r}
library(knitr)
library(rmarkdown)
library(tidyverse)
library(anytime)
library(scales)
source.data <- read.csv(params$filename, stringsAsFactors = FALSE)

occupancy<-as_tibble(source.data)
occupancy

#dataset is large, create sample 20% to  test
smp<-sample_frac(occupancy, 0.02)
smp

linmod = lm(OCCUPANCY~SECTOR, data=occupancy)

summary(linmod)

linmod = lm(OCCUPANCY~months.Date(as.Date(OCCUPANCY_DATE)), data=occupancy)

ggplot(smp, aes(as.Date(OCCUPANCY_DATE), OCCUPANCY)) + 
                  geom_point()+theme_light() +
                  labs(x="DATE", y="OCCUPANCY", title="DATE and Occupancy") + 
                  stat_smooth(method="lm", se=FALSE) +
                  scale_x_date(labels=date_format("%B %Y"), breaks=date_breaks("3 months"))
summary(linmod)

```

##Conclusion of your analysis if any in the .RMD file.
```{r}
cat("The analized data set includes figures showing users (homeless peole) who entered the shelter system from January 2020 over a 12-month period \n
The impacts of job losses amid the economic slowdown caused by COVID-19 is the couse of increased number of people entering shelters \n
The chosen dataset is very clean and provides good quality data \n
     1. Analysisi shows that Etobicoke area has the lest shelter in Toronto and Old Toronto has the hiest number of shelters\n
     2. Shelter by sectors or types: Families, Men, Women, and Co-ed \n
     3. The data shows that shelters for Men have the highest capacity, then follow family shelters, then women’s shelters, and finally Co-ed. \n
     4. The data shows thet in some periods shelters occupied over the Shelter Capacity \n\n
     https://github.com/sprotas-hash/data-analysis.git \n \n
     I was not able to created the PDf file using 'Knit to PDF', all the time it gives me error, but I was able to create HTML file using 'Knit to HTML' and I downloaded it to git")
```